Seagrass mapping and monitoring \\ along the coasts of Crete, Greece

### 2011-02-11 MScDiplom_Lemenkova.tex
written '1.8. Research Assumptions'p.11

	%\acro{BRDF}{Bidirectional Reflectance Distribution Function}

@@

\begin{figure}[H]
	\centering
	\includegraphics[scale=0.48]{UML_ClassDiagram.jpg}
	\caption{General methodological approach for the analysis of spectral signatures. \href{http://argouml.tigris.org/}{ArgoUML}, \href{http://www.gimp.org/}{Gimp}. \\ Page: $<$\pageref{page-9}$>$}
	\label{fig:1.10}
\end{figure}

The spectral reflectance, distinct for each seafloor cover type, enables their discrimination on the image. The pre-processing
of the images includes imagery corrections for atmospheric noises and effects of the water
column. Reflectance spectra of the seagrass canopy at different depths of the water-column are
analysed for the discrimination of their spectral signatures, enabling to separate various seafloor types
during classification. The results of the of imagery classification are analysed for the detection of the
dynamics in \textit{P.oceanica} seagrass distribution along the northern coasts of Crete. Aerial imagery from
Google Earth with high spatial resolution, suitable for the large-scale detailed mapping of seagrass
mattes, is used for the improvement of the accuracy of large seagrass meadows and separate mattes
within the meadows. The \textit{in situ} underwater videometric measurements of the seafloor are collected
during the fieldwork in Crete, for the validation of the classification results and to determine the exact
current distribution of the \textit{P.oceanica} meadows. The image processing includes steps of the remote
sensing techniques, i.e. calibration, masking from land and cloud, atmospheric correction, sea surface
glint and depth effects correction as recommended \cite{Matarrese08}\label{Matarrese08}. During the image classification
working step the training sites for the supervised classification methods are designed, as well as its
control and trials of different classification approaches (Unsupervised, K-means or Isodata;
Supervised, Maximum Likelihood).

\subsection{Research assumptions}\index[concepts]{Research!assumptions}
The wind roughening the water surface causes sun glitter, and the values spectral reflectance may contain disturbances and noisy if the solar zenith angle is too high, say approaching 60-70 degrees. Therefore, the measurements should be made at midday (noon) with as low zenith angle (Fig.\ref{fig:A.2-1})\label{zenith} as possible.
\pagebreak

@@

\subsection[Application of the remote sensing...]{Application of the remote sensing data towards seagrass mapping}\index[concepts]{Remote sensing}
Various methods and approaches have been applied towards mapping of the seagrasses, based on
digitized aerial photographs, \ac{GPS} data, remote sensing and \ac{SCUBA}-based (Fig.\ref{fig:A.52}) fieldwork measurements.
\ac{SCUBA}-based \textit{in-situ} observations, though providing high resolution and accuracy results in seagrass
mapping, is limited in application, because of their time consumption, weather-dependency and
unsuitability for the case of monitoring large areas of water for small-scale mapping. 
The underwater videography\index[concepts]{Remote sensing!underwater videography} with a \ac{GPS}\index[concepts]{Remote sensing!GPS} is a tool of seagrass monitoring which has certain advantages, i.e.
high spatial and visual resolution, non-destructive sampling, effectiveness at all depths and rapid data
collection in the field \cite{Schultz08}\label{Schultz08}.

\begin{wrapfigure}{r}{0.4\textwidth}
	\centering
	\includegraphics[scale=0.35]{Fig-11.jpg}
	\caption{\textit{In situ} optical reflectance spectra of seagrass. Shaded areas - \% of spectra lying within the range of reflectance. 
		White lines - mean spectra. Source: \cite{Hochberg03b}\label{Hochberg03b}. Page: $<$\pageref{page-21}$>$}
	\label{fig:13}
\end{wrapfigure}

@@

Studies of spectral reflectance\index[concepts]{Seawater optical properties!spectral reflectance} of the different seagrass species comparing to the spectra of sand and other seafloor cover types \cite{Vahtmae06}\label{Vahtmae06} prove that spectra of green, brown and red benthic
macroalgae differ from each other, as well as from sand and deep water reflectance spectra. These
differences are well detectable by the means of the remote sensing research methods. Comparing to
the terrestrial plants, aquatic vegetation, \textit{inter alia} seagrass, cannot be detected using red edge of the
spectrum, as these wavelengths are significantly absorbed by water \cite{Kirk94}\label{Kirk94}, as well as by
scattering and absorption by phytoplankton. Some authors \cite{Dierssen03}\label{Dierssen03} also use
spectrally based radiative transfer approach to quantitatively estimate shallow-water bathymetry and
leaf area index (\ac{LAI}) of the seagrass. The spectral reflectance in general is the
result of the spectral absorption in different bands, typical for each target object.
Spectral reflectance\index[concepts]{Seawater optical properties!spectral reflectance} of the seagrass  (Fig.\ref{fig:13}) is largely influenced by the water
depth where it is located, and is generally decreasing in values by increasing depths.
The most important spectral diapason\index[concepts]{Seawater optical properties!spectral reflectance!spectral diapason} for marine mapping of submerged vegetation and, particularly, for seagrass, is 350-800 nm \cite{Barille09}\label{Barille09}.
Using airborne imagery for retrospective data (before 1970s) together with the most
recent imagery allows to detect changes in seagrass distribution on over different years
and to analyse dynamics of the seagrass distribution \cite{Ball09}\label{Ball09}. \\
Another important advantage of the application of the remote sensing data for mapping of submerged aquatic vegetation has
commercial nature: using remote sensing data and methods enables more low-cost and up-to-date
seagrass mapping \cite{Mumby99}\label{Mumby99}, and is especially useful for the areas where the fieldwork data
capturing is unavailable.
Seagrass meadows may reach spatial scales from several up to hundreds of metres, therefore they are
susceptible by the means of satellite imagery from remote sensors, both with moderate resolution
(e.g., \href{http://landsat.gsfc.nasa.gov/}{Landsat MSS, Landsat TM, Landsat ETM+}, \href{http://envisat.esa.int/instruments/meris/}{MERIS} from \ac{ESA}, \href{http://asterweb.jpl.nasa.gov/}{ASTER}, \href{http://modis.gsfc.nasa.gov/}{MODIS}) and high resolution as
well (e.g., \href{http://www.satimagingcorp.com/gallery-ikonos.html}{IKONOS}, \href{http://www.digitalglobe.com/index.php/85/QuickBird}{Quickbird}, \href{http://www.spotimage.fr/}{SPOT}, \href{http://www.itres.com/products/imagers/casi550}{CASI}). The possibility of their application towards seagrass mapping varies and is limited by the technical characteristics (Table \ref{tab:1}) and resolution of these sensors  (Fig.\ref{fig:2.3}).
In the next paragraphs (\nameref{sec:2.4.1} and \nameref{sec:2.4.2}) we will briefly discuss research experience of the application of satellite imagery towards seagrass mapping, its advantages and limitations, reported by various researchers.

@@

%%%%%%%%%%%%%%%%%  Section  3  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Materials and methods}\label{sec:3}
\counterwithin{figure}{section}
\counterwithin{table}{section}

@@

\begin{figure}[H]
	\centering
	\subfloat [iPAQ, HP]{\label{fig:3.3a}\includegraphics[width=0.2\textwidth]{Fig-19a.jpg}}
	\subfloat [Waterproof Otterbox for iPAQ]{\label{fig:3.3b}\includegraphics[width=0.2\textwidth]{Fig-19b.jpg}}
	\subfloat [GPS]{\label{fig:3.3c}\includegraphics[width=0.15\textwidth]{Fig-19c.jpg}}
	\subfloat [Olympus waterproof camera]{\label{fig:3.3d}\includegraphics[width=0.2\textwidth]{Fig-19d.jpg}}
	\caption{Fieldwork equipment. Pages: $<$\pageref{page-15}$>$, $<$\pageref{page-24}$>$, $<$\pageref{page-25}$>$}
	\label{fig:3.3}
\end{figure}

@@

\begin{figure}[H]
	\centering
	\subfloat [view from below]{\label{fig:22a}\includegraphics[width=0.3\textwidth]{Fig-20a.jpg}}
	\subfloat [view \textit{en-face}]{\label{fig:22a}\includegraphics[width=0.3\textwidth]{Fig-20b.jpg}}
	\subfloat [view from above]{\label{fig:22a}\includegraphics[width=0.3\textwidth]{Fig-20c.jpg}}
	\caption{Scheme of placements of the Olimpus cameras during the measurements. Page: $<$\pageref{page-26}$>$}
	\label{fig:3.5}
\end{figure}

The videographic measurements and photos, enabling to detect types of the
seafloor, were captured by means of three underwater digital video cameras \ac{Olympus ST} 8000  (Fig.\ref{fig:3.3d})\label{page-26}, at
five minutes interval along the tracklog path (Fig.\ref{fig:3.6}).
research sampling design also included measurements of depths (Fig. \ref{fig:A.4}), because bathymetry is one of the
most determining factors for the seagrass locations. Measurements of depths were performed during
the fieldwork using cord, \ac{iPAQ}, and markers in order to assure that the videometric measurements
are taken at depths not more than four meters, in the shelf area.
As a result, nine transects were established of one m wide and 20 m long track, to cover seagrass beds
with videographic measurements. The \ac{GPS} allowed to capture measurement locations on the \ac{iPAQ},
encapsulated in a plastic waterproof Otterbox (Fig \ref{fig:3.3b}). The camera were adjusted horizontally by a leveler and
mounted under the bottom of the boat Fig.\ref{fig:A.51}\label{page-27} to capture video footage and imagery (Fig.\ref{fig:3.5}). The data
were taken at proper weather conditions: sunny, serene and cloud-free days with glassy sea state. 
The locations of the route were randomly selected in the areas of the Ligaria beach, to ensure most dense
coverage of the seagrass meadows in the research area. The underwater measurements of the seagrass
coverage were carried out by taking video footage and photos of ca 0.5 m2 size each (Fig.\ref{fig:A.7}, Fig.\ref{fig:A.55}).
The results of the underwater videometric measurements include series of digital images helping to
classify seafloor cover types (Fig.\ref{fig:A.56}) and seagrass meadows, according to the differences in the structure,
colour, texture and shapes of the depicted objects.

\begin{figure}[H]
	\centering
	\includegraphics[scale=0.50]{Fig-21.jpg}
	\caption{Locations of the video measurements and GPS tracklogs, Ligaria. Page: $<$\pageref{page-25}$>$}
	\label{fig:3.6}
\end{figure}

Along the Ligaria beach there are several types of the seagrass landscapes, namely, seagrass meadows
continuously covering vast areas (Fig.\ref{fig:A.3} (a)), aggregated seagrass patches, represented by separate mattes with short irregular channels between them (Fig.\ref{fig:A.3} (b)) and isolated seagrass patches, or mattes,
which located separately from broad meadows (Fig.\ref{fig:A.3} (c), Fig.\ref{fig:A.56}).
The results of the underwater videometric measurements of the \ac{Olympus ST} cameras made during the
ship route include nine total tracklog routes in the selected research area, including series of
consequent images, completely covering the area under the boat path. The received data contain
information on seagrass presence within the study area, distribution of seagrass \textit{P.oceanica} meadows
and nature of the seafloor cover types: rocks, sandy, mixed (Fig. \ref{fig:A5}). Seagrass species on the Ligaria and Agia
Pelagia beach consist of 	extit{P.oceanica}. The types of sediment on the Ligaria vary from coarse sand,
\textit{P.oceanica} patches, sand, mud, rock gravel and fine sand (Fig.\ref{fig:A.6}).

@@

A collection of visible spectra of two seafloor cover types - \textit{P.oceanica} and carbonate sand - consist of multiple measurement sets of \textit{P.oceanica} made by means of RAMSES hyperspectral radiometer: 1-350 for 15th October (fragment on Tab. \ref{tab:A.8}), 1-445 for 14th October for 	\textit{P.oceanica}; 1-106 for water without sediments, measured in aquarium tank, 1-27 for seawater with sediments measured in aquarium tank; 1-75 for carbonate sand (Tab. \ref{tab:17}). All these data have been analyzed and statistically tested. We measured mean, extreme values (min-max), median and statistical quartiles, in order to visualize distribution of the values at various spectral wavelengths. Carbonate sand was measured at wavelengths 402-750 nm, while \textit{P.oceanica} - at the interval 318-951 nm. We visualized the behavior of different spectra on relevant graphs (Fig. \ref{fig:4.4}, \ref{fig:A.35}, \ref{fig:A.28}). 

@@

calculation of the median, mean, quartiles and other statistical values at every data set (see \nameref{app}, Fig. \ref{fig:A.25}, Tab. \ref{tab:17}).

@@

\subsection{Review of the collected data}
The remote sensing imagery was collected for the Crete island area covering the research period and
area, enabling the field data to be used for the calibration and validation. The available aerial and
satellite imagery are commonly used for the mapping of seagrass landscapes and their application is
proven by various research papers. The imagery includes satellite multi-spectral imagery (Landsat-TM,
ETM+) and aerial imagery from the Google Earth.
The overview of the collected data enables to summarize their following types:
\begin{itemize}
	\item[] 1. Optical spectra of \textit{P.oceanica}, carbonate sand, seawater with sediments and seawater measured in aquarium tank, without sediments, at different environmental conditions (e.g., Fig. \ref{fig:A.23})
	\item[]2. Aerial imagery from the \href{http://www.google.com/earth/index.html}{Google Earth}
	\item[]3. Satellite images from various open sources (e.g. Landsat, Tab \ref{tab:20}) (previews: Fig. \ref{fig:A.37})
	\item[]4. Results of underwater videometric measurements of the \ac{Olympus ST} cameras made during the ship route
\end{itemize}
The available satellite imagery (Fig.\ref{fig:A.38}) and aerial images are read into the ArcGIS project.
The available broadband and hyperspectral remote sensing data are used for the mapping of the
seagrass in shelf areas no deeper than 4.0 meters, for the environmental monitoring in order to detect
the dynamics in changes of the seagrass distribution along Crete during the past 10 years using
different satellite images for 2000-2010.

@@

%
\subsection{Analysis of spectral signatures}
The distinguishing spectral signatures for various seafloor types (e.g. seagrass species, coral reefs,
various types of sand, mud, other sediments, (Fig.\ref{fig:4.4})) exist in well-defined and narrow (10–20 nm)
wavelength ranges. 
The values of their spectral reflectance are accepted as constant. The results of
spectral measurements enable to analyze, whether \textit{P.oceanica} is spectrally distinct from other sea
floor types with changing environmental conditions, using the differences in their spectral signatures
on the graphs in a \ac{WASI} software. \\The \ac{WASI}, a
software tool for analyzing and simulating the most common types of spectra \cite{Gege05}\label{Gege05}, is highly
suitable for the seagrass spectral analysis.
There are several environmental characteristics, included in \ac{WASI} interface (Fig.\ref{fig:4.7}), which influence the
results of water spectral reflectance, e.g. different bottom depths, concentration of suspended particles
in water column, water temperature, sun angle, concentration of Gelbstoff (\ac{CDOM}), concentration of phytoplankton, aerosol scattering, exponent of backscattering by small
particle \cite{Gege04}\label{Gege04}.

\begin{figure}[H]
	\centering
	\includegraphics[scale=0.35]{GNU-20-new.jpg}
	\caption{Statistical comparison of spectral reflectance of \textit{P.oceanica} and carbonate sand. Selected measurements sets. \href{http://www.gnuplot.info/}{Gnuplot} graph. Page: $<$\pageref{page-47}$>$}
	\label{fig:28}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[scale=0.5]{GNU-17.jpg}
	\caption{Optical properties of the sea water with sediments, measured in aquarium tank. Agia Pelagia district, Crete. \href{http://www.gnuplot.info/}				{Gnuplot}. Page: $<$\pageref{page-21}$>$}\label{fig:4.2}
\end{figure}

>>>>

	\acro{BRDF}{Bidirectional Reflectance Distribution Function}

@@

\begin{figure}[H]
	\centering
	\includegraphics[scale=0.63]{UML_ClassDiagram.jpg}
	\caption{General methodological approach for the analysis of spectral signatures. \href{http://argouml.tigris.org/}{ArgoUML}, \href{http://www.gimp.org/}{Gimp}. \\ Page: $<$\pageref{page-9}$>$}
	\label{fig:1.10}
\end{figure}

The spectral reflectance, distinct for each seafloor cover type, enables their discrimination on the image. The pre-processing
of the images includes imagery corrections for atmospheric noises and effects of the water
column. Reflectance spectra of the seagrass canopy at different depths of the water-column are
analysed for the discrimination of their spectral signatures, enabling to separate various seafloor types
during classification. The results of the of imagery classification are analysed for the detection of the
dynamics in \textit{P.oceanica} seagrass distribution along the northern coasts of Crete. Aerial imagery from
Google Earth with high spatial resolution, suitable for the large-scale detailed mapping of seagrass
mattes, is used for the improvement of the accuracy of large seagrass meadows and separate mattes
within the meadows. The \textit{in situ} underwater videometric measurements of the seafloor are collected
during the fieldwork in Crete, for the validation of the classification results and to determine the exact
current distribution of the \textit{P.oceanica} meadows. The image processing includes steps of the remote
sensing techniques, i.e. calibration, masking from land and cloud, atmospheric correction, sea surface
glint and depth effects correction as recommended \cite{Matarrese08}\label{Matarrese08}. During the image classification
working step the training sites for the supervised classification methods are designed, as well as its
control and trials of different classification approaches (Unsupervised, K-means or Isodata;
Supervised, Maximum Likelihood).
\pagebreak

\subsection{Research assumptions}\index[concepts]{Research!assumptions}
The general research assumption that is used for further applications of the remote sensing methods of this work is that optical properties of the spectral reflectance of the seawater with different bottom substrata enable to discriminate seafloor cover types using remote sensing products from broadband sensors.\vspace{1ex}\\
The accepted model-specific parameters of the \ac{WASI} simulator are described in Table 4.2. The weather conditions for the measurements are assumed to be perfect: clear, sunny, windless days, because wind roughen water surface causing sun glitter, therefore, values of spectral reflectance may contain disturbances. \\ Sky radiance might be influenced by multiple reflectance between the sea surface albedo and the atmosphere and in general, sky radiance increases rapidly while viewing zenith angle is near to $90\,^{\circ}$ (i.e. in the evening hours). Therefore, ideally measurements should be done at noon, with as low zenith angle (Fig.\ref{fig:A.2-1})\label{zenith} as possible; otherwise, if the solar zenith angle is too high (e.g. approaching $60-70\,^{\circ}$), received data may contain noise. For \ac{WASI} simulations reflection factor of sky radiance is taken as 0.0201 with simulated ideal conditions: viewing angle ($0\,^{\circ}$ = nadir). The interval of water temperature taken as default lies in the diapason $17-25\,^{\circ}$ as to simulate the conditions of the Mediterranean Sea. However, in real time conditions the sun zenith angle of $45\,^{\circ}$ has been accepted as suitable.\vspace{1ex}\\
The anisotropy factor of upwelling radiation or the quality (Q-) factor, showing the directionally dependency of the radiance, is taken as 5. We accept some values of optical parameters as default values at \ac{WASI}. Thus, concentration of phytoplankton is accepted at the interval of $0.035-0.089 mg^{-1}$ and concentration of large suspended particles is given to 8. Reference wavelength for \ac{CDOM} (Gelbstoff) absorption is equal to 440. The coefficient of attenuation remains equal to 1.0546, as set up by default at \ac{WASI}.\vspace{1ex}\\
We also assume that concentration of non-chlorophyll particles (absorption at 0) as well as concentration of small suspended particles is equal to zero, so we do not count them in this work. Exponent of \ac{CDOM} (Gelbstoff) absorption is accepted as 0.0140. Finally, the \ac{BRDF} of bottom reflectance (sand), which defines the reflection of light  is at an opaque surface (Fig.A.2.5), is assumed to be 0.318 $sr^{-1}$.

@@

\subsection[Application of the remote sensing...]{Application of the remote sensing data towards seagrass mapping}\index[concepts]{Remote sensing}\label{page-15}
Various methods and approaches have been applied towards mapping of the seagrasses, based on
digitized aerial photographs, \ac{GPS} data, remote sensing and \ac{SCUBA}-based (Fig.\ref{fig:A.52}) fieldwork measurements.
\ac{SCUBA}-based \textit{in-situ} observations, though providing high resolution and accuracy results in seagrass
mapping, is limited in application, because of their time consumption, weather-dependency and
unsuitability for the case of monitoring large areas of water for small-scale mapping. 
The underwater videography\index[concepts]{Remote sensing!underwater videography} with a \ac{GPS}\index[concepts]{Remote sensing!GPS} is a tool of seagrass monitoring which has certain advantages, i.e.
high spatial and visual resolution, non-destructive sampling, effectiveness at all depths and rapid data
collection in the field \cite{Schultz08}\label{Schultz08}.

\begin{wrapfigure}{r}{0.4\textwidth}
	\centering
	\includegraphics[scale=0.35]{Fig-11.jpg}
	\caption{\textit{In situ} optical reflectance spectra of seagrass. Shaded areas - \% of spectra lying within the range of reflectance. 
		White lines - mean spectra. Source: \cite{Hochberg03b}\label{Hochberg03b}. \\ Pages: $<$\pageref{page-17}$>$, $<$\pageref{page-21}$>$}
	\label{fig:13}
\end{wrapfigure}

@@

Studies of spectral reflectance\index[concepts]{Seawater optical properties!spectral reflectance} of the different seagrass species comparing to the spectra of sand and other seafloor cover types \cite{Vahtmae06}\label{Vahtmae06} prove that spectra of green, brown and red benthic
macroalgae differ from each other, as well as from sand and deep water reflectance spectra. These
differences are well detectable by the means of the remote sensing research methods. Comparing to
the terrestrial plants, aquatic vegetation, \textit{inter alia} seagrass, cannot be detected using red edge of the
spectrum, as these wavelengths are significantly absorbed by water \cite{Kirk94}\label{Kirk94}, as well as by
scattering and absorption by phytoplankton. Some authors \cite{Dierssen03}\label{Dierssen03} also use
spectrally based radiative transfer approach to quantitatively estimate shallow-water bathymetry and
leaf area index (\ac{LAI}) of the seagrass. The spectral reflectance in general is the
result of the spectral absorption in different bands, typical for each target object.
Spectral reflectance\index[concepts]{Seawater optical properties!spectral reflectance} of the seagrass  (Fig.\ref{fig:13})\label{page-17} is largely influenced by the water
depth where it is located, and is generally decreasing in values by increasing depths.
The most important spectral diapason\index[concepts]{Seawater optical properties!spectral reflectance!spectral diapason} for marine mapping of submerged vegetation and, particularly, for seagrass, is 350-800 nm \cite{Barille09}\label{Barille09}.
Using airborne imagery for retrospective data (before 1970s) together with the most
recent imagery allows to detect changes in seagrass distribution on over different years
and to analyse dynamics of the seagrass distribution \cite{Ball09}\label{Ball09}. \\
Another important advantage of the application of the remote sensing data for mapping of submerged aquatic vegetation has
commercial nature: using remote sensing data and methods enables more low-cost and up-to-date
seagrass mapping \cite{Mumby99}\label{Mumby99}, and is especially useful for the areas where the fieldwork data
capturing is unavailable.
Seagrass meadows may reach spatial scales from several up to hundreds of metres, therefore they are
susceptible by the means of satellite imagery from remote sensors, both with moderate resolution
(e.g., \href{http://landsat.gsfc.nasa.gov/}{Landsat MSS, Landsat TM, Landsat ETM+}, \href{http://envisat.esa.int/instruments/meris/}{MERIS} from \ac{ESA}, \href{http://asterweb.jpl.nasa.gov/}{ASTER}, \href{http://modis.gsfc.nasa.gov/}{MODIS}) and high resolution as
well (e.g., \href{http://www.satimagingcorp.com/gallery-ikonos.html}{IKONOS}, \href{http://www.digitalglobe.com/index.php/85/QuickBird}{Quickbird}, \href{http://www.spotimage.fr/}{SPOT}, \href{http://www.itres.com/products/imagers/casi550}{CASI}). The possibility of their application towards seagrass mapping varies and is limited by the technical characteristics (Table \ref{tab:1}) and resolution of these sensors  (Fig.\ref{fig:2.3}).
In the next paragraphs (\nameref{sec:2.4.1} and \nameref{sec:2.4.2}) we will briefly discuss research experience of the application of satellite imagery towards seagrass mapping, its advantages and limitations, reported by various researchers.

@@

%%%%%%%%%%%%%%%%%  Section  3  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Materials and methods}\label{sec:3}\label{page-22}
\counterwithin{figure}{section}
\counterwithin{table}{section}

@@

\begin{figure}[H]
	\centering
	\subfloat [iPAQ, HP]{\label{fig:3.3a}\includegraphics[width=0.2\textwidth]{Fig-19a.jpg}}
	\subfloat [Waterproof Otterbox for iPAQ]{\label{fig:3.3b}\includegraphics[width=0.2\textwidth]{Fig-19b.jpg}}
	\subfloat [GPS]{\label{fig:3.3c}\includegraphics[width=0.15\textwidth]{Fig-19c.jpg}}
	\subfloat [Olympus waterproof camera]{\label{fig:3.3d}\includegraphics[width=0.2\textwidth]{Fig-19d.jpg}}
	\caption{Fieldwork equipment. Pages: $<$\pageref{page-15}$>$, $<$\pageref{page-24}$>$, $<$\pageref{page-25}$>$, $<$\pageref{page-26}$>$, $<$\pageref{page-27}$>$}
	\label{fig:3.3}
\end{figure}

@@

\begin{figure}[H]
	\centering
	\subfloat [view from below]{\label{fig:22a}\includegraphics[width=0.3\textwidth]{Fig-20a.jpg}}
	\subfloat [view \textit{en-face}]{\label{fig:22a}\includegraphics[width=0.3\textwidth]{Fig-20b.jpg}}
	\subfloat [view from above]{\label{fig:22a}\includegraphics[width=0.3\textwidth]{Fig-20c.jpg}}
	\caption{Scheme of placements of the Olimpus cameras during the measurements. Pages: $<$\pageref{page-26}$>$, $<$\pageref{page-27}$>$}
	\label{fig:3.5}
\end{figure}

The videographic measurements and photos, enabling to detect types of the
seafloor, were captured by means of three underwater digital video cameras \ac{Olympus ST} 8000  (Fig.\ref{fig:3.3d})\label{page-26}, at
five minutes interval along the tracklog path (Fig.\ref{fig:3.6}).
research sampling design also included measurements of depths (Fig. \ref{fig:A.4}), because bathymetry is one of the
most determining factors for the seagrass locations. Measurements of depths were performed during
the fieldwork using cord, \ac{iPAQ}, and markers in order to assure that the videometric measurements
are taken at depths not more than four meters, in the shelf area.
As a result, nine transects were established of one m wide and 20 m long track, to cover seagrass beds
with videographic measurements. The \ac{GPS} allowed to capture measurement locations on the \ac{iPAQ},
encapsulated in a plastic waterproof Otterbox (Fig \ref{fig:3.3b}). The camera were adjusted horizontally by a leveler and
mounted under the bottom of the boat Fig.\ref{fig:A.51}\label{page-27} to capture video footage and imagery (Fig.\ref{fig:3.5}). The data
were taken at proper weather conditions: sunny, serene and cloud-free days with glassy sea state. 
The locations of the route were randomly selected in the areas of the Ligaria beach, to ensure most dense
coverage of the seagrass meadows in the research area. The underwater measurements of the seagrass
coverage were carried out by taking video footage and photos of ca 0.5 m2 size each (Fig.\ref{fig:A.7}, Fig.\ref{fig:A.55}).
The results of the underwater videometric measurements include series of digital images helping to
classify seafloor cover types (Fig.\ref{fig:A.56}) and seagrass meadows, according to the differences in the structure,
colour, texture and shapes of the depicted objects\label{page-27}.

\begin{figure}[H]
	\centering
	\includegraphics[scale=0.50]{Fig-21.jpg}
	\caption{Locations of the video measurements and GPS tracklogs, Ligaria. Pages: $<$\pageref{page-25}$>$, $<$\pageref{page-26}$>$}
	\label{fig:3.6}
\end{figure}

Along the Ligaria beach there are several types of the seagrass landscapes, namely, seagrass meadows
continuously covering vast areas (Fig.\ref{fig:A.3} (a)), aggregated seagrass patches, represented by separate mattes with short irregular channels between them (Fig.\ref{fig:A.3} (b)) and isolated seagrass patches, or mattes,
which located separately from broad meadows (Fig.\ref{fig:A.3} (c), Fig.\ref{fig:A.56}).
The results of the underwater videometric measurements of the \ac{Olympus ST} cameras made during the
ship route include nine total tracklog routes in the selected research area, including series of
consequent images, completely covering the area under the boat path. The received data contain
information on seagrass presence within the study area, distribution of seagrass \textit{P.oceanica} meadows
and nature of the seafloor cover types: rocks, sandy, mixed (Fig.\ref{fig:A.6}, Fig.\ref{fig:A.5}, Fig.\ref{fig:A.3}). Seagrass species on the Ligaria and Agia Pelagia beach consist of \textit{P.oceanica}. The types of sediment on the Ligaria vary from coarse sand,
\textit{P.oceanica} patches, sand, mud, rock gravel and fine sand (Fig.\ref{fig:A.6}).

@@

A collection of visible spectra of two seafloor cover types - \textit{P.oceanica} and carbonate sand - consist of multiple measurement sets of \textit{P.oceanica} made by means of RAMSES hyperspectral radiometer: 1-350 for 15th October (fragment on Tab. \ref{tab:A.8}), 1-445 for 14th October for \textit{P.oceanica}; 1-106 for water without sediments, measured in aquarium tank, 1-27 for seawater with sediments measured in aquarium tank; 1-75 for carbonate sand (Tab.\ref{tab:17}). All these data have been analyzed and statistically tested. We measured mean, extreme values (min-max), median and statistical quartiles, in order to visualize distribution of the values at various spectral wavelengths. Carbonate sand was measured at wavelengths 402-750 nm, while \textit{P.oceanica} - at the interval 318-951 nm. We visualized the behavior of different spectra on relevant graphs (Fig. \ref{fig:4.4}, \ref{fig:A.35}, \ref{fig:A.28}).\label{page-28}

@@

calculation of the median, mean, quartiles and other statistical values at every data set (see \nameref{app},\label{dataprep} Fig. \ref{fig:A.25}, Tab. \ref{tab:17}).

@@

\subsection{Review of the collected data}
The remote sensing imagery was collected for the Crete island area covering the research period and
area, enabling the field data to be used for the calibration and validation. The available aerial and
satellite imagery are commonly used for the mapping of seagrass landscapes and their application is
proven by various research papers. The imagery includes satellite multi-spectral imagery (Landsat-TM,
ETM+) and aerial imagery from the Google Earth.
The overview of the collected data enables to summarize their following types:
\begin{itemize}
	\item[] 1. Optical spectra of \textit{P.oceanica}, carbonate sand, seawater with sediments and seawater measured in aquarium tank, without sediments, at different environmental conditions (e.g., Fig. \ref{fig:A.23})
	\item[]2. Aerial imagery from the \href{http://www.google.com/earth/index.html}{Google Earth}
	\item[]3. Satellite images from various open sources (e.g. Landsat, Tab \ref{tab:20}) (previews: Fig. \ref{fig:A.37})
	\item[]4. Results of underwater videometric measurements of the \ac{Olympus ST} cameras made during the ship route
\end{itemize}
The available satellite imagery (Fig.\ref{fig:A.38}) \label{page-30}and aerial images are read into the ArcGIS project.
The available broadband and hyperspectral remote sensing data are used for the mapping of the
seagrass in shelf areas no deeper than 4.0 meters, for the environmental monitoring in order to detect
the dynamics in changes of the seagrass distribution along Crete during the past 10 years using
different satellite images for 2000-2010.

@@

%
\subsection{Analysis of spectral signatures}
The distinguishing spectral signatures for various seafloor types (e.g. seagrass species, coral reefs,
various types of sand, mud, other sediments, (Fig.\ref{fig:4.4})\label{page-31} exist in well-defined and narrow (10–20 nm)
wavelength ranges. 
The values of their spectral reflectance are accepted as constant. The results of
spectral measurements enable to analyze, whether \textit{P.oceanica} is spectrally distinct from other sea
floor types with changing environmental conditions, using the differences in their spectral signatures
on the graphs in a \ac{WASI} software. \\The \ac{WASI}, a
software tool for analyzing and simulating the most common types of spectra \cite{Gege05}\label{Gege05}, is highly
suitable for the seagrass spectral analysis.
There are several environmental characteristics, included in \ac{WASI} interface (Fig.\ref{fig:4.7}), which influence the
results of water spectral reflectance, e.g. different bottom depths, concentration of suspended particles
in water column, water temperature, sun angle, concentration of Gelbstoff (\ac{CDOM}), concentration of phytoplankton, aerosol scattering, exponent of backscattering by small
particle \cite{Gege04}\label{Gege04}.

\begin{figure}[H]
	\centering
	\includegraphics[scale=0.35]{GNU-20-new.jpg}
	\caption{Statistical comparison of spectral reflectance of \textit{P.oceanica} and carbonate sand. Selected measurements sets. \href{http://www.gnuplot.info/}{Gnuplot} graph. Pages: $<$\pageref{page-28}$>$, $<$\pageref{page-47}$>$, $<$\pageref{dataprep}$>$}
	\label{fig:28}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[scale=0.5]{GNU-17.jpg}
	\caption{Optical properties of the sea water with sediments, measured in aquarium tank. Agia Pelagia district, Crete. \href{http://www.gnuplot.info/}				{Gnuplot}. Pages: $<$\pageref{page-21}$>$, $<$\pageref{page-31}$>$}\label{fig:4.2}
\end{figure}

The backward-scattering coefficient (bb), also included in \ac{WASI}, is a
fundamental optical property which plays a central role in the ocean-color remote sensing, providing
the remotely sensed optical signal, as well as suspended particle distributions, water clarity, and
underwater visibility \cite{Maffione97}\label{Maffione97}. \ac{WASI} enables simulation of backscattering of pure
water, large and small particles. The values of all these parameters can be redacted and changed
manually. However, the most important, major factors affecting the \textit{in-situ} conditions are water depth (Fig.\ref{fig:4.9})\label{page-32}
and cover fraction of the seafloor types: \textit{P.oceanica} and carbonate sand.
Seagrass \textit{P.oceanica} can be mapped using remotely acquired spectral imagery, but only if it has distinctive
reflectance signatures at different depths. Therefore, the depths of the shelf area are the first variable
condition for this research question. The depths values chosen for the current research are lesser than
3.5 meters, covering shelf zone, and providing the best environmental conditions for the seagrass
\textit{P.oceanica}: 0.5, 2.0 and 3.5 meters with an interval of 1.5 meters. 
The 3.5 m depth limit was chosen based on the analysis of the separability of seagrass reflectance signatures, received by the means of
previous \textit{in-situ} measurements (year 2009) of the radiance and irradiance of water in Agia Pelagia
bay, which indicated that \textit{P.oceanica} seagrass could be only well separated within depths of 3.5 m.
A statistical analysis of \ac{WASI}-simulated spectral reflectance has been used in order to answer the first
research question: whether the \textit{P.oceanica} spectra is spectrally distinct at varying environmental \textit{in-situ}
conditions, and if \textit{P.oceanica} remains spectrally distinct with the increasing water column depth.
To answer this research question, different seafloor cover types are discriminated using data of the
broadband remote sensing.

\begin{figure}[H]
	\centering
	\includegraphics[scale=0.3]{GNU-18-legend.png}
	\caption{Multiplot showing spectral reflectance of the seawater with sediments, measured in aquarium tank, Agia Pelagia district, Crete. \href{http://www.gnuplot.info/}{Gnuplot}. Two complimentary graphs below show the results of the statistical analysis.\\ Pages: $<$\pageref{page-28}$>$, $<$\pageref{dataprep}$>$, $<$\pageref{page-31}$>$}
	\label{fig:4.4}
\end{figure}
